Experience Report: Haskell in Computational Biology

### 2012-03-08 mrfy_experience_report.tex
haskell types!

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
\remark{Time to introduce the elements of Viterbi's equations!}
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model.
Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.

We employed a variety of Haskell data types to represent the hidden Markov
model.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as \texttt{Int}.
Similarly, to avoid a great deal of data transformation, each state's emission
and insertion probabilities were simply
\begin{verbatim}type EProbs = V.Vector Double \end{verbatim} while the states themselves
were algebraic datatypes: 
\begin{verbatim}data HMMState = Match | Insert | Delete \end{verbatim}

\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}

% time for types.
% first, ideal: amino acids as an algebraic DT
% next, back off this: we're working with an existing input format
% so, how is the graph represented (vector of state-transitions, 
% vector of emissions)
% so show those types

@@

%% break the viterbi types up and insert into the Plan7 explanation.
%% go over this with NR; I can't make it flow.

\begin{figure}
\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
\end{figure}

@@

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}I_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}I_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

>>>>

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence of amino acids.
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
Mathematically, we represent a query sequence as
$\left\{x_{1}...x_{n}\right\}$.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as 
\begin{verbatim}
type AA = Int
\end{verbatim}
and thus our query sequence as
\begin{verbatim}
type querySeq = Vector AA
\end{verbatim}

The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{verbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{verbatim}
to represent the seven allowed state transitions.

Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
query sequence element $x_{i}$.
In Haskell, we represent a node as a group of transition probabilities
\begin{verbatim}
data TProbs = TProbs { m_m :: TProb
                     , m_i :: TProb
                     , m_d :: TProb
                     , i_m :: TProb
                     , i_i :: TProb
                     , d_m :: TProb
                     , d_d :: TProb
                     }
\end{verbatim}
and a vector of emission probabilities for the match and insertion states.
\begin{verbatim}
type EProbs = V.Vector logProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{verbatim}

@@

%% break the viterbi types up and insert into the Plan7 explanation.
%% go over this with NR; I can't make it flow.
% 
% \begin{figure}
% \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
% \end{figure}
%

@@

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}
